Computing Direct Sum Decompositions

### 2025-04-27 paper/main.tex
paper edits before final push

\def\cF{\mathcal F}
\def\OO{\mathcal O}
\def\FF{\mathbb F}
\def\PP{\mathbb P}
\def\ZZ{\mathbb Z}
\makeop{Gr}
\makeop{Bl}
\makeop{Eff}

@@

\begin{abstract}
  We describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory. We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards (including in the non-graded case) and syzygies over Artinian rings.
\end{abstract}

@@

The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields. Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter). For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
The need for and absence of an algorithm to do so is noted, for example, in
\cite[15.10.9]{Eisenbud95}.

@@

Let $R$ be a $\N$-graded ring with $R_0 = k$ a field. Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a general degree-zero endomorphism $\phi \in [\End M]_0$. Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as an $\mu(M)\times\mu(M)$ matrix with entries in $R$. We will refer to $\chi(\phi) = \det(\phi-\lambda\id_M)$ as the ``characteristic polynomial of $\phi$'', which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions in $\lambda$ only in the algebraic closure of the total ring of fractions of $R$.

\begin{prop}\label{prop:eigenvalues}
  The solutions of the characteristic polynomial $\chi(\phi)$ are in $\bar k$.
\end{prop}
\begin{proof}
  Since $M$ is $\N$-graded, assume its generators are ordered with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \begin{equation}\label{eq:blocks}
    \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
    \end{pmatrix},
  \end{equation}
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. %In particular, $B_i$ is an $k$-endomorphism on the $k$-vector space spanned by all generators of degree $d_i$.

  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id)$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in the algebraic closure of $k$.
\end{proof}

We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''. Further, let $\mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$.

@@

Note that as long as each $\psi_i$ in \Cref{lem:decomposition} is nontrivial (not zero or identity endomorphisms), the kernel summands will be nontrivial, however, they may not necessarily be indecomposable. Also note that if the $\psi_i$ are complementary surjections, $\im\psi$ will be zero.

\begin{prop}\label{prop:split-surj}
  Let $\lambda_1,\dots,\lambda_r\in k$ be the eigenvalues of $\phi$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu(\lambda_j)}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
\end{prop}
\begin{proof}
  First we show that each $\psi_j$ is a split surjection $M\to\im\psi_j$, hence $M = \ker\psi_j \oplus \im\psi_j$. To see this, note that $R$-linear row and column operations on endomorphisms of $M$ correspond to pre- and post-composition by $R$-linear maps on $M$. In particular, performing row and column operations only involving elements of $R_0 = k$ on $\phi$ we may transform each $B_i$ into Jordan canonical form
  \begin{equation}
    J_i = \begin{pmatrix}
      J_{i,1} & 0       & \dots  & 0      \\
      0       & J_{i,2} & \dots  & 0      \\
      \vdots  & \vdots  & \ddots & \vdots \\
      0       & 0       & \dots  & J_{i,t},
    \end{pmatrix}
  \end{equation}
  where each Jordan block has entries over $k$ and a fixed eigenvalue. Suppose $J_{i,j}$ is the block with eigenvalue $\lambda_j$, then $N_i = J_{i,j} - \lambda_j$ is a nilpotent matrix whose size is the multiplicity of $\lambda-\lambda_j$ in $\chi(B_i)$. Since $\mu(\lambda_j)$ is the sum of the multiplicities of $\lambda_j$ in each $B_i$, the matrix $N_i^{\mu(\lambda_j)}$ vanishes, hence $(J_i-\lambda_j)^{\mu(\lambda_j)}$ is a matrix with some 1s along the diagonal and 0s elsewhere. This implies that there are auto morphisms $\alpha$ and $\beta$ of $M$ (corresponding to column operations clearing the entries above and to the right of each nonzero diagonal) such that $\alpha \circ (\phi - \lambda_j\id_M)^{\mu(\lambda_j)} \circ \beta: M\to M$ is represented by a diagonal matrix $\psi'_j$ with 1s and 0s along the diagonal, which is a split surjection.

  Finally, if $\lambda_j$ and $\lambda_k$ are different then the nilpotent blocks in $J_i-\lambda_j$ and $J_i-\lambda_k$ are distinct, therefore $\ker\psi_j \cap \ker\psi_k = 0$, thus by \Cref{lem:decomposition} we have the desired decomposition.
\end{proof}

@@

Note that \Cref{alg:graded} is characteristic independent. In particular, since over a large enough field the eigenvalues of a general endomorphism $\phi$ are distinct, the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable, thus \Cref{alg:graded} will produce indecomposable summands in one non-recursive iteration. We will see via \Cref{lem:term}\mahrud{or lem:distinct?} that this will result in an indecomposable decomposition of $M$ over any field.

@@

We begin with the observation that if $M$ is a finitely generated $R$-module and $A\in\End_R(M)$ is an idempotent, then $M$ decomposes as $\im A \oplus \coker A$. If $A$ is neither an isomorphism nor the zero morphism, then both factors are nonzero and $M$ is decomposable. Note that $A$ also acts on the $k$-vector space $M/\m M$.

@@

\begin{lem}\label{lem:distinct}
  Retain the notation of \Cref{lem:term}.
  \begin{itemize}
  \item If $R$ is graded and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (necessarily in $k$)
  \item If $R$ is local and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
  \end{itemize}
\end{lem}

@@

\begin{rem}
  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs all the eigenvalues of the endomorphism $A$ of $M/\m M$ to be contained in $k$. While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field. However, the general linear combinations we take in Step~\ref{item:End0} of \Cref{alg:graded} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
  See \Cref{ex:elliptic} for a demonstration of the necessity of extending the base field.
\end{rem}

>>>>

\def\cF{\mathcal F}
\def\OO{\mathcal O}
\def\FF{\mathbb F}
\def\PP{\mathbb P}
\def\ZZ{\mathbb Z}
\makeop{Gr}
\makeop{Bl}
\makeop{Eff}
\def\wpsi{\widehat\psi}

@@

\begin{abstract}
  We describe and prove correctness of two practical algorithms for finding indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$. The first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory. We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards (including in the non-graded case) and syzygies over Artinian rings.
\end{abstract}

@@

The problems of finding isomorphism classes of indecomposable modules with a given property, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra and representation theory and even in areas such as topological data analysis. Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter). For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
The need for and absence of an algorithm to do so is noted, for example, in
\cite[15.10.9]{Eisenbud95}.

@@

Let $R$ be a $\N$-graded ring with $R_0 = k$ a field. Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a general degree-zero endomorphism $\phi \in [\End M]_0$. Fix a set of minimal homogeneous generators $m_1,\dots,m_{\mu(M)}$ of $M$ so that $\phi$ may be presented as a $\mu(M)\times\mu(M)$ matrix with entries in $R$. We will refer to $\chi_\phi(\lambda) = \det(\phi-\lambda\id_M)$ as the ``characteristic polynomial of $\phi$'', which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.

\begin{prop}\label{prop:eigenvalues}
  Let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi\in[\End M]_0$. The solutions of the characteristic polynomial $\chi_\phi(\lambda)$ are the same as the eigenvalues of $A$, thus they lie in the algebraic closure $\bar k$.
\end{prop}
\begin{proof}
  Since $M$ is $\N$-graded, assume its generators are ordered with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \begin{equation}\label{eq:blocks}
    \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
    \end{pmatrix},
  \end{equation}
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $\m$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. %In particular, $B_i$ is an $k$-endomorphism on the $k$-vector space spanned by all generators of degree $d_i$.

  Then the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id) = \det(\phi - \lambda \id_{M/\m M})$ is a univariate polynomial in $\lambda$ with coefficients in $k$, thus has solutions in $\bar k$.
\end{proof}

We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''. Further, let $\mu_j \coloneqq \mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$. Note that the multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$.

@@

Note that as long as each $\psi_i$ in \Cref{lem:decomposition} is nontrivial (not zero or identity endomorphisms), the kernel summands will be nontrivial, however, they may not necessarily be indecomposable. Also note that $\im\psi$ may be zero.

\pagebreak
\begin{prop}\label{prop:split-surj}
  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi$ over $k$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu_j}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
\end{prop}
\begin{proof}
  First, suppose $\lambda_{r+1},\dots,\lambda_t$ are the other eigenvalues of $\phi$ over the algebraic closure $\bar k$ and set $\rho = \psi_{r+1}\circ\dots\circ\psi_t$, then by the Cayley--Hamilton theorem $\chi_\phi(\phi) = \psi_1\circ\dots\circ\psi_r\circ\rho = \psi\circ\rho = 0$. Since the factors in $\chi_\phi(\phi)$ can be reordered, we can write $0 = \psi\circ\rho = \rho \circ \psi$. In particular, suppose the desired direct sum decomposition existed over $\bar k$, then we have \mahrud{WHY??}
  \begin{align*}
    M = \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus(\ker\psi_{r+1}\oplus\dots\oplus\ker\psi_t)
    &= \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus\ker\rho \\
    &= \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus\im\psi.
  \end{align*}
  Thus it suffices to prove the direct sum decomposition only over the algebraic closure.

  If $r = 1$ then $\im\psi_1=0$ is zero, hence we get the trivial decomposition $M = M \oplus 0$.

  %  We will find a right inverse to $\psi_1$ and $\psi_2$ using $\chi_\phi(\phi) = 0$
  If $r = 2$, let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi$ and $A_i\coloneqq(A - \lambda_i)^{\mu_i}$ the map induced by $\psi_i$ for $i=1,2$. Observe that
  \begin{align*}
  \chi_\phi(\phi) = (\phi - \lambda_1\id_M)^{\mu_1} (\phi - \lambda_2\id_M)^{\mu_2}
  &= \left(\phi^{\mu_1} - \dots + c_1\phi - c_0\id_M\right)\circ\psi_2 \\
  &= \left(\phi^{\mu_1} - \dots + c_1\phi\right)\circ\psi_2 - c_0\psi_2 = 0,
  \end{align*}
  where $c_0 = -(-\lambda_1)^{\mu_1}$ and $c_1 = (-\lambda_1)^{\mu_1-1}$, and therefore we have a left pseudoinverse
  \[
  \psi_2
  = c_0^{-1} \left(\phi^{\mu_2} - \dots + c_1\phi \right) \circ\psi_2
  % = \phi\circ c_0^{-1} \left(\phi^{\mu_2-1} - \dots + c_1 \right) \circ\psi_2.
  \]
  This means we have a short exact sequence $0 \to \im\psi_2 \hookrightarrow M \to \coker\psi_2 \to 0$ which is split by the map $c_0^{-1} \left(\phi^{\mu_2} - \dots + c_1\phi \right)$, % which is to say $M \to \im\psi_2$ is a split surjection.

  \begin{align*}
  \chi_\phi(\phi) = (\phi - \lambda_2\id_M)^{\mu_2} (\phi - \lambda_1\id_M)^{\mu_1}
  &= \psi_2\circ\left(\phi^{\mu_1} - \dots + c_1\phi - c_0\id_M\right) \\
  &= \psi_2 c_0 - \psi_2\circ \left(\phi^{\mu_1} - \dots + c_1\phi\right) = 0,
  \end{align*}

  % $\lambda_1^{-1}A_1$ is an idempotent and by \Cref{lem:idemp} (and scaling by $\lambda_1$ for simplicity) we get $M = \ker\psi_1 \oplus \im\psi_1$.
% \devlin{But $\lambda_1\inv A_1$ is not guaranteed to be an idempotent: e.g., if $\lambda_1 = 1$ with multiplicity 2, then $A_1$ has the block $\begin{pmatrix}1 & 2 \\ 0 & 1\end{pmatrix}$ (i.e., a $2\times 2$ Jordan block squared), and it's not an idempotent.}

  Otherwise, observe that $\wpsi_j\coloneqq \psi_1\circ\dots\circ\psi_{j-1}\circ\psi_{j+1}\circ\dots\circ\psi_r $ has only one nonzero eigenvalue over $k$, namely $\widehat\lambda_j \coloneqq \prod_{i\neq j}(\lambda_j-\lambda_i)$: Using \Cref{prop:eigenvalues}, eigenvalues of $\phi$ are the same as the eigenvalues of $A$. Say $A v = \lambda_\ell v$ for $v\in M/\m M$ and apply the $k$-linear map induced by $\psi_1\circ\dots\circ\psi_r$ to get
  \begin{align}\label{eq:eigenvalues-psi}
    \prod(A-\lambda_i)v
    &= (A - \lambda_1)\cdots(A - \lambda_{r-1})(\lambda_\ell - \lambda_r)v \nonumber \\
    &= (A - \lambda_1)\cdots(\lambda_\ell - \lambda_{r-1})(\lambda_\ell - \lambda_r)v
    = \dots = \prod(\lambda_\ell - \lambda_i)v,
  \end{align}
  where the product is over $i=1,\dots,r$. In particular, since the $\lambda_i$ are distict the $k$-linear map $\widehat A_j\coloneqq (A - \lambda_1)\dots(A-\lambda_{j-1})(A - \lambda_{j+1})\dots(A - \lambda_r)$ induced by $\wpsi_j$ removes the $(\lambda_\ell-\lambda_j)$ factor from \eqref{eq:eigenvalues-psi}, hence the eigenvalue of $\widehat A_j$ corresponding to $v$ is $\prod_{i\neq j}(\lambda_j-\lambda_i)$ if $\lambda_\ell=\lambda_j$ and 0 otherwise, with multiplicities $\mu_j$ and $\mu(M) - \mu_j$ respectively.

  In particular, $\sfrac{\widehat A_j}{\widehat\lambda_j}$ is an idempotent, so by \Cref{lem:idemp} (and scaling by $\widehat\lambda_j$ for simplicity) we have $M = \ker\wpsi_j \oplus \im\wpsi_j$ and further each $\psi_j$ is a split surjection $M\to\im\psi_j = \ker\wpsi_j$. Finally, note that elements of $\ker\psi_i = \im\wpsi_i$ must be generated by columns of $\phi$ corresponding to the eigenvalue $\lambda_i$, hence $\ker\psi_i \cap \ker\psi_j = 0$ for $i\neq j$, thus by \Cref{lem:decomposition} we have the desired decomposition.
\end{proof}

%% \begin{proof}
%%   First we show that each $\psi_j$ is a split surjection $M\to\im\psi_j$, hence $M = \ker\psi_j \oplus \im\psi_j$. To see this, note that $R$-linear row and column operations on endomorphisms of $M$ correspond to pre- and post-composition by $R$-linear maps on $M$. In particular, performing row and column operations only involving elements of $R_0 = k$ on $\phi$ we may transform each $B_i$ into Jordan canonical form
%%   \begin{equation}
%%     J_i = \begin{pmatrix}
%%       J_{i,1} & 0       & \dots  & 0      \\
%%       0       & J_{i,2} & \dots  & 0      \\
%%       \vdots  & \vdots  & \ddots & \vdots \\
%%       0       & 0       & \dots  & J_{i,t},
%%     \end{pmatrix}
%%   \end{equation}
%%   where each Jordan block has entries over $k$ and a fixed eigenvalue. Suppose $J_{i,j}$ is the block with eigenvalue $\lambda_j$, then $N_i = J_{i,j} - \lambda_j$ is a nilpotent matrix whose size is the multiplicity of $\lambda-\lambda_j$ in $\chi_{B_i}(\lambda)$. Since $\mu_j$ is the sum of the multiplicities of $\lambda_j$ in each $B_i$, the matrix $N_i^{\mu_j}$ vanishes, hence $(J_i-\lambda_j)^{\mu_j}$ is a matrix with some 1s along the diagonal and 0s elsewhere. This implies that there are auto morphisms $\alpha$ and $\beta$ of $M$ (corresponding to column operations clearing the entries above and to the right of each nonzero diagonal) such that $\alpha \circ (\phi - \lambda_j\id_M)^{\mu_j} \circ \beta: M\to M$ is represented by a diagonal matrix $\psi'_j$ with 1s and 0s along the diagonal, which is a split surjection.

%%   Finally, if $\lambda_j$ and $\lambda_k$ are different then the nilpotent blocks in $J_i-\lambda_j$ and $J_i-\lambda_k$ are distinct, therefore $\ker\psi_j \cap \ker\psi_k = 0$, 
%% \end{proof}

@@

We will see via \Cref{lem:distinct} that this will result in an indecomposable decomposition of $M$ over any field.

\begin{rem}
Note that \Cref{alg:graded} does not depend on the characteristic of the field.
Since over a large enough field (i.e., when $|k|$ is larger than the number of summands) the eigenvalues of a general endomorphism $\phi$ are distinct (by a variant of the argument of \Cref{lem:term}),
the summands $\ker\psi_i$ and $\im\psi$ will be indecomposable; thus \Cref{alg:graded} will produce indecomposable summands in one non-recursive iteration. 
When $|k|$ is less than or approximately the number of summands, the algorithm will be recursive.
\end{rem}

@@

We begin with the observation that if $M$ is a finitely generated $R$-module and $A\in\End_R(M)$ is an idempotent, then $M$ decomposes as $\im A \oplus \coker A$. If $A$ is neither an isomorphism nor the zero morphism, then both factors are nonzero and $M$ is decomposable.

Note that $A$ also acts on the $k$-vector space $M/\m M$.

@@

\begin{lem}\label{lem:distinct}
  Retain the notation of \Cref{lem:term}.
  \begin{itemize}
  \item If $R$ is graded and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (in $\bar k$)
  \item If $R$ is local and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
  \end{itemize}
\end{lem}

@@

\begin{rem}
  Note that 
\Cref{alg:graded} and \Cref{alg:local} are 
 quite sensitive to the ground field $k$, because they needs all the eigenvalues of the endomorphism 
%$A$ of $M/\m M$ 
to be contained in $k$. While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field. However, the general linear combinations we take in Step~\ref{item:End0} of \Cref{alg:graded} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
  See \Cref{ex:elliptic} for a demonstration of the necessity of extending the base field.

We also note that the algorithm detects the necessity of a field extension: if the characteristic polynomial of the endomorphism does not factor over $k$, the algorithm will calculate the splitting field of $k$; the user can then opt to extend the ground field to this splitting field.
\end{rem}
